---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rumenGP

<!-- badges: start -->
<!-- badges: end -->

**rumenGP** is an R package for processing and modeling *in vitro* rumen gas production data generated by systems such as the ANKOM RF Gas Production System.

The package currently provides tools for:

- Importing ANKOM RF data
- Parsing ANKOM incubation timestamps
- Converting pressure measurements to gas production
- Joining and validating experimental metadata
- Fitting the Zwietering-modified Gompertz model
- Calculating kinetic parameters and goodness-of-fit statistics
- Flagging potentially problematic bottles
- Plotting raw gas production curves and model fits
- Documenting bottle exclusions for reproducible analyses

## Installation

Development version:

```r
# install.packages("devtools")

devtools::install_github(
  "YOUR_GITHUB_USERNAME/rumenGP"
)
```

Or install from a local source:

```r
devtools::install_local(
  "path/to/rumenGP"
)
```

## Example workflow

```{r}
library(rumenGP)
```

### Load packaged example files

```{r}
files <- example_data()

files
```

### Import ANKOM data

```{r}
raw_data <- read_ankom(
  files$ankom
)

metadata <- read_metadata(
  files$metadata
)
```

### Validate metadata

```{r}
metadata <- validate_metadata(
  metadata
)
```

### Process ANKOM data

```{r}
gp <- process_ankom(
  raw_data,
  metadata,
  headspace_ml = 210,
  temperature_c = 39,
  zero_negative_pressure = TRUE
)
```

### Validate processed data

```{r}
gp <- validate_ankom(
  gp
)
```

### Fit Gompertz model

```{r}
gompertz_fit <- fit_gompertz(
  gp
)
```

### Summarize fit results

```{r}
summary(gompertz_fit)
```

### Flag potentially problematic bottles

```{r}
flags <- flag_gompertz(
  gompertz_fit
)

flags
```

### Plot a raw gas production curve

```{r, eval = FALSE}
plot_gp(
  gp,
  head = "1"
)
```

### Plot a Gompertz fit

```{r, eval = FALSE}
plot_gompertz_fit(
  gompertz_fit,
  head = "1"
)
```

### Exclude problematic bottles

```{r}
gp_clean <- exclude_heads(
  gp,
  heads = c("10"),
  reason = "Sensor malfunction"
)
```

## Current development status

### Data handling

- ✅ `read_ankom()`
- ✅ `read_metadata()`
- ✅ `parse_ankom_time()`
- ✅ `process_ankom()`

### Validation

- ✅ `validate_metadata()`
- ✅ `validate_ankom()`

### Gompertz modeling

- ✅ `fit_gompertz()`
- ✅ `flag_gompertz()`
- ✅ `plot_gp()`
- ✅ `plot_gompertz_fit()`
- ✅ `plot_residuals()`
- ✅ `exclude_heads()`

### Planned features

- ⏳ Logistic model
- ⏳ Mitscherlich model
- ⏳ France model
- ⏳ Michaelis–Menten model
- ⏳ Model comparison framework
- ⏳ Automatic model selection
- ⏳ Publication-ready reports

## Citation

If you use **rumenGP** in research, please cite:

```text
Rodrigues, A. A. & Mantovani, H. C.
rumenGP: Tools for modelling in vitro rumen gas production data.
```

(Citation information will be updated as the package develops.)
