
---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rumenGP

<!-- badges: start -->
<!-- badges: end -->

**rumenGP** is an R package for processing and modeling *in vitro* rumen gas production data generated by systems such as the ANKOM RF Gas Production System.

The package provides a reproducible workflow for:

- Importing ANKOM RF output files
- Parsing ANKOM incubation timestamps
- Converting pressure measurements into gas production
- Validating metadata and ANKOM datasets
- Fitting multiple gas production models
- Comparing competing models
- Detecting problematic bottles
- Visualizing model fits and residuals
- Generating treatment-level summaries

---

## Installation

Development version:

```r
# install.packages("devtools")

devtools::install_github(
  "araujorodrig-lab/rumenGP"
)
```

Or install from a local source:

```r
devtools::install_local(
  "path/to/rumenGP"
)
```

---

## Example workflow

```{r}
library(rumenGP)
```

### Load packaged example files

The package includes a small example dataset for learning the workflow and testing package functionality.

```{r}
files <- example_data()
```

### Import ANKOM data

```{r}
raw_data <- read_ankom(
  files$ankom
)

metadata <- read_metadata(
  files$metadata
)
```

### Validate metadata

```{r}
metadata <- validate_metadata(
  metadata
)
```

### Process ANKOM data

```{r}
gp <- process_ankom(
  raw_data,
  metadata,
  headspace_ml = 210,
  temperature_c = 39,
  zero_negative_pressure = TRUE
)
```

### Validate processed data

```{r}
gp <- validate_ankom(
  gp
)
```

### Fit models

```{r}
gompertz_fit <- fit_gompertz(gp)

logistic_fit <- fit_logistic(gp)

mitscherlich_fit <- fit_mitscherlich(gp)

mm_fit <- fit_mm(gp)
```

### Summaries

```{r}
summary(gompertz_fit)

summary(logistic_fit)

summary(mitscherlich_fit)

summary(mm_fit)
```

### Flag potentially problematic fits

```{r}
flags <- flag_model(
  gompertz_fit
)

head(flags)
```

### Compare models

```{r}
comparison <- compare_models(
  Gompertz = gompertz_fit,
  Logistic = logistic_fit,
  Mitscherlich = mitscherlich_fit,
  MichaelisMenten = mm_fit
)

comparison
```

### Rank models

```{r}
rank_models(
  comparison
)
```

### Determine the best model by treatment

```{r}
treatment_comparison <- compare_models_by_treatment(
  Gompertz = gompertz_fit,
  Logistic = logistic_fit,
  Mitscherlich = mitscherlich_fit,
  MichaelisMenten = mm_fit
)

ranked_treatments <- rank_models_by_treatment(
  treatment_comparison
)

best_model_by_treatment(
  ranked_treatments
)
```

---

## Visualization

### Raw gas production curve

```{r, eval = FALSE}
plot_gp(
  gp,
  head = "1"
)
```

### Plot model fit

```{r, eval = FALSE}
plot_fit(
  gompertz_fit,
  head = "1"
)
```

### Plot residuals

```{r, eval = FALSE}
plot_residuals(
  gompertz_fit,
  head = "1"
)
```

### Compare residuals across models

```{r, eval = FALSE}
plot_residual_comparison(
  Gompertz = gompertz_fit,
  Logistic = logistic_fit,
  Mitscherlich = mitscherlich_fit,
  MichaelisMenten = mm_fit,
  head = "1"
)
```

### Treatment mean comparison

```{r, eval = FALSE}
plot_treatment_mean(
  Gompertz = gompertz_fit,
  Logistic = logistic_fit,
  Mitscherlich = mitscherlich_fit,
  MichaelisMenten = mm_fit,
  treatment = "Plant_A"
)
```

### All treatment means

```{r, eval = FALSE}
plot_all_treatment_means(
  Gompertz = gompertz_fit,
  Logistic = logistic_fit,
  Mitscherlich = mitscherlich_fit,
  MichaelisMenten = mm_fit
)
```

---

## Quality Control Workflow

The recommended workflow is:

```text
Import data
    ↓
Validate metadata
    ↓
Process ANKOM data
    ↓
Validate processed data
    ↓
Fit models
    ↓
Flag problematic bottles
    ↓
Visual inspection
    ↓
Document exclusions
    ↓
Refit models
    ↓
Compare models
```

Example:

```{r, eval = FALSE}
gp_clean <- exclude_heads(
  gp,
  heads = c("10"),
  reason = "Sensor malfunction"
)
```

---

## Implemented Models

Currently available:

- Gompertz
- Logistic
- Mitscherlich
- Michaelis-Menten

---

## Current Development Status

### Data Handling

- ✅ `read_ankom()`
- ✅ `read_metadata()`
- ✅ `parse_ankom_time()`
- ✅ `process_ankom()`

### Validation

- ✅ `validate_metadata()`
- ✅ `validate_ankom()`

### Modeling

- ✅ `fit_gompertz()`
- ✅ `fit_logistic()`
- ✅ `fit_mitscherlich()`
- ✅ `fit_mm()`

### Diagnostics and QC

- ✅ `flag_model()`
- ✅ `exclude_heads()`
- ✅ `plot_residuals()`
- ✅ `plot_residual_comparison()`

### Visualization

- ✅ `plot_gp()`
- ✅ `plot_fit()`
- ✅ `plot_all_fits()`
- ✅ `plot_treatment_mean()`
- ✅ `plot_all_treatment_means()`
- ✅ `plot_model_comparison_treatment()`

### Model Comparison

- ✅ `compare_models()`
- ✅ `rank_models()`
- ✅ `compare_models_by_treatment()`
- ✅ `rank_models_by_treatment()`
- ✅ `best_model_by_treatment()`

### Planned Features

- ⏳ Logistic-Exponential model
- ⏳ Additional exponential models
- ⏳ Automated model selection
- ⏳ Biological plausibility scoring
- ⏳ Treatment reporting utilities
- ⏳ Mixed-effects workflow
- ⏳ pkgdown website

---

## Citation

If you use **rumenGP** in research, please cite:

```text
Rodrigues, A. A. & Mantovani, H. C.

rumenGP:
An R package for processing and modeling
in vitro rumen gas production data.
```

(Citation information will be updated as the package develops.)
