
---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rumenGP

<!-- badges: start -->
<!-- badges: end -->

**rumenGP** is an R package for importing, processing,
visualizing, fitting, comparing, and interpreting
*in vitro* rumen gas production data.

The package supports:

- ANKOM RF datasets
- Manually collected gas-volume datasets
- Pressure-based datasets (PSI or kPa)
- Twelve built-in kinetic models
- User-defined kinetic models
- Model comparison and ranking
- Treatment-level model evaluation
- Diagnostic and visualization tools

---

## Installation

Development version:

```r
# install.packages("devtools")

devtools::install_github(
  "araujorodrig-lab/rumenGP"
)
```

Or install from a local source:

```r
devtools::install_local(
  "path/to/rumenGP"
)
```

---

## Package Workflow

```text
ANKOM RF
        ↓
process_ankom()
        ↓
     rumen_gp

Manual Gas Volume
        ↓
as_rumen_gp()
        ↓
     rumen_gp

Pressure Data
        ↓
as_rumen_gp()
        ↓
pressure_to_volume()
        ↓
     rumen_gp

             ↓

 Built-in Models
          +
  Custom Models

             ↓

 Comparison
 Ranking
 Diagnostics
 Visualization
```

---

## Example Workflow: ANKOM Data

```{r}
library(rumenGP)
```

### Load and Import Example Data

The package includes a small example dataset that can be used
to learn the workflow and test package functionality.

```{r}
files <- example_data()

raw_data <- read_ankom(
  files$ankom
)

metadata <- read_metadata(
  files$metadata
)
```

### Inspect ANKOM Data

The ANKOM RF file contains pressure measurements collected
during the incubation period.

```{r}
head(raw_data)
```

### Inspect Metadata

The metadata table contains bottle identifiers,
treatments, and replicate information.

```{r}
head(metadata)
```

### Validate metadata

```{r}
metadata <- validate_metadata(
  metadata
)
```

### Process ANKOM data

```{r}
gp <- process_ankom(
  raw_data,
  metadata,
  headspace_ml = 210,
  temperature_c = 39,
  zero_negative_pressure = TRUE
)
```

### Validate processed data

```{r}
gp <- validate_ankom(
  gp
)
```

---

## Example Workflow: Manual Gas Volume Data

```{r}
manual_volume <- data.frame(

  Bottle = c(
    1,1,1,
    2,2,2
  ),

  Treatment = c(
    "Control","Control","Control",
    "Corn","Corn","Corn"
  ),

  Time = c(
    0,4,8,
    0,4,8
  ),

  Gas = c(
    0,20,40,
    0,35,60
  )

)

gp_manual <- as_rumen_gp(
  data = manual_volume,
  head_col = "Bottle",
  treatment_col = "Treatment",
  time_col = "Time",
  gas_col = "Gas"
)

head(gp_manual)
```

---

## Example Workflow: Pressure Data

```{r}
manual_pressure <- data.frame(

  Bottle = rep(
    1,
    10
  ),

  Time = c(
    0,2,4,6,8,
    12,16,24,36,48
  ),

  PSI = c(
    0,
    0.2,
    0.5,
    0.8,
    1.2,
    1.8,
    2.5,
    3.2,
    4.0,
    4.5
  )

)

gp_pressure <- as_rumen_gp(
  data = manual_pressure,
  head_col = "Bottle",
  time_col = "Time",
  pressure_col = "PSI",
  pressure_unit = "psi",
  headspace_volume = 60
)

head(gp_pressure)
```

Pressure measurements are automatically converted
to gas volume using the same conversion framework
used by `process_ankom()`.

---

## Model Fitting

```{r}
groot_fit <- fit_groot(gp)

brody_fit <- fit_brody(gp)

gompertz_fit <- fit_gompertz(gp)
```

### Model summaries

```{r}
summary(groot_fit)
```

---

## Compare Models

```{r}
comparison <- compare_models(

  Groot = fit_groot(gp),

  Brody = fit_brody(gp),

  Gompertz = fit_gompertz(gp)

)

comparison
```

### Rank Models

```{r}
rank_models(
  comparison
)
```

---

## Treatment-Level Comparison

```{r}
treatment_comparison <-
  compare_models_by_treatment(

    Groot = fit_groot(gp),

    Brody = fit_brody(gp),

    Gompertz = fit_gompertz(gp)

  )

ranked_treatments <-
  rank_models_by_treatment(
    treatment_comparison
  )

best_model_by_treatment(
  ranked_treatments
)
```

---

## User-Defined Models

Researchers can fit their own kinetic equations
using `fit_custom()`.

```{r}
custom_fit <- fit_custom(

  data = gp,

  formula =
    Gas_mL ~
      A *
      (
        Time_h /
        (
          Time_h + K
        )
      ),

  start = list(
    A = 150,
    K = 10
  ),

  lower = c(
    A = 0,
    K = 0
  ),

  model_name = "Hyperbolic"

)

summary(custom_fit)
```

Compare custom and built-in models:

```{r}
compare_models(

  Groot = fit_groot(gp),

  Brody = fit_brody(gp),

  Hyperbolic = custom_fit

)
```

---

## Visualization

### Raw gas production

```{r, eval = FALSE}
plot_gp(
  gp,
  head = "1"
)
```

### Plot model fit

```{r, eval = FALSE}
plot_fit(
  groot_fit,
  head = "1"
)
```

### Plot residuals

```{r, eval = FALSE}
plot_residuals(
  groot_fit,
  head = "1"
)
```

### Treatment means

```{r, eval = FALSE}
plot_treatment_mean(
  Groot = fit_groot(gp),
  Brody = fit_brody(gp),
  treatment = "Control"
)
```

---

## Quality Control Workflow

```text
Import data
    ↓
Validate metadata
    ↓
Create rumen_gp object
    ↓
Fit models
    ↓
Flag problematic bottles
    ↓
Inspect residuals
    ↓
Document exclusions
    ↓
Refit models
    ↓
Compare models
```

Example:

```{r, eval = FALSE}
gp_clean <- exclude_heads(
  gp,
  heads = c("10"),
  reason = "Sensor malfunction"
)
```

---

## Implemented Models

### Built-in Models

- Brody
- Dual Logistic
- EXP0
- EXPL
- Gompertz
- Groot
- LE0
- LEL
- Logistic
- Mitscherlich
- Michaelis-Menten
- Orskov and McDonald
- Burr XII
- Inverse Paralogistic

### Model Equivalence

Several model formulations included in rumenGP are
mathematically equivalent and are provided because
different scientific disciplines and publications
prefer different parameterizations.

#### Groot ↔ Generalized Michaelis-Menten

The Groot model

V(t) = VF / [1 + (b/t)^k]

is mathematically equivalent to the generalized
Michaelis-Menten model

V(t) = A t^c / (t^c + K^c)

Parameter correspondence:

- VF = A
- b = K
- k = c

Both formulations produce identical fitted values,
residuals, model diagnostics, AIC, BIC, RMSE,
and R-squared when convergence is achieved.

Researchers may choose either formulation according
to the terminology commonly used in their field.

#### Groot ↔ Michaelis-Menten ↔ Log-logistic

The Log-logistic model

V(t) = VF(rt)^a / [1 + (rt)^a]

can be rewritten as

V(t) = VF t^a / [t^a + (1/r)^a]

which is mathematically identical to the Groot
and generalized Michaelis-Menten formulations.

Parameter correspondence:

| Groot | Michaelis-Menten | Log-logistic |
|---------|---------|---------|
| VF | A | VF |
| b | K | 1/r |
| k | c | a |

These formulations describe the same underlying
curve and differ only in parameterization.

Consequently, rumenGP already represents the
Log-logistic model through the existing Groot
and generalized Michaelis-Menten implementations.

### Recent Model Additions

#### Burr XII

Equation:

V(t) = VF [1 - (1 + (rt)^a)^(-p)]

Parameters:

- VF: asymptotic gas production
- r: rate parameter
- a: shape parameter
- p: shape parameter

Advantages:

- Highly flexible sigmoidal behavior
- Accommodates diverse fermentation profiles
- Often provides strong goodness-of-fit
- Useful for comparative model evaluation

Limitations:

- Four-parameter model
- Greater risk of overfitting than simpler models
- Parameter interpretation may be less intuitive

#### Inverse Paralogistic

Equation:

V(t) = VF [1 + (rt)^(-a)]^(-a)

Parameters:

- VF: asymptotic gas production
- r: rate parameter
- a: shape parameter

Advantages:

- Flexible sigmoidal behavior
- Biologically meaningful asymptote
- Performs well across diverse curve shapes

Limitations:

- Shape parameter may be difficult to interpret
- Requires positive incubation times
- Less commonly used than traditional rumen models

Neither Burr XII nor Inverse Paralogistic should
be regarded as universally superior.

Model performance depends on feed type,
experimental design, data quality, and the
model-selection criteria used.

---

## Recommended Model Selection Workflow

No single gas production model is expected to
perform best for every dataset.

A recommended workflow is:

1. Fit several biologically plausible models.
2. Verify model convergence.
3. Inspect fitted curves visually.
4. Examine residual patterns.
5. Compare RMSE, AIC, and BIC.
6. Evaluate parameter plausibility.
7. Select the model most appropriate for the
   research objective.

Published comparative work has reported strong
performance for Burr XII, Inverse Paralogistic,
and Log-logistic formulations across diverse
feed datasets.

Because the Log-logistic formulation is
mathematically equivalent to both Groot and
generalized Michaelis-Menten models, rumenGP
already provides this curve family through
those existing parameterizations.

---

## Major Features

### Data Import

- `process_ankom()`
- `as_rumen_gp()`

### Modeling

- 15 built-in kinetic models
- Burr XII
- Inverse Paralogistic
- `fit_custom()`

### Diagnostics

- `summary()`
- `flag_model()`
- `exclude_heads()`

### Visualization

- `plot_gp()`
- `plot_fit()`
- `plot_residuals()`
- `plot_treatment_mean()`

### Model Comparison

- `compare_models()`
- `rank_models()`
- `compare_models_by_treatment()`
- `rank_models_by_treatment()`
- `best_model_by_treatment()`
- `model_win_frequency()`

---

### Completed

- ✅ ANKOM RF workflow
- ✅ Generic data import
- ✅ Pressure-based workflows
- ✅ Fifteen kinetic models
- ✅ Burr XII model
- ✅ Inverse Paralogistic model
- ✅ Custom model framework
- ✅ Model comparison framework
- ✅ Treatment-level model ranking
- ✅ Model-equivalence documentation
- ✅ Diagnostic workflows
- ✅ Visualization tools
- ✅ Five package vignettes
- ✅ pkgdown website
- ✅ Clean R CMD check

### Development Status

rumenGP is currently maintained and developed
as an open-source framework for rumen gas
production analysis.

Current releases support:

- ANKOM RF workflows
- Manual gas production workflows
- Pressure-based workflows
- Fifteen nonlinear gas production models
- User-defined kinetic models
- Model comparison and ranking
- Treatment-level evaluation
- Visualization and diagnostics

Future releases may include additional
reporting, model-evaluation, and statistical
workflows.

---
