
---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rumenGP

<!-- badges: start -->
<!-- badges: end -->

**rumenGP** is an R package for importing, processing,
visualizing, fitting, comparing, and interpreting
*in vitro* rumen gas production data.

The package supports:

- ANKOM RF datasets
- Manually collected gas-volume datasets
- Pressure-based datasets (PSI or kPa)
- Twelve built-in kinetic models
- User-defined kinetic models
- Model comparison and ranking
- Treatment-level model evaluation
- Diagnostic and visualization tools

---

## Installation

Development version:

```r
# install.packages("devtools")

devtools::install_github(
  "araujorodrig-lab/rumenGP"
)
```

Or install from a local source:

```r
devtools::install_local(
  "path/to/rumenGP"
)
```

---

## Package Workflow

```text
ANKOM RF
        ↓
process_ankom()
        ↓
     rumen_gp

Manual Gas Volume
        ↓
as_rumen_gp()
        ↓
     rumen_gp

Pressure Data
        ↓
as_rumen_gp()
        ↓
pressure_to_volume()
        ↓
     rumen_gp

             ↓

 Built-in Models
          +
  Custom Models

             ↓

 Comparison
 Ranking
 Diagnostics
 Visualization
```

---

## Example Workflow: ANKOM Data

```{r}
library(rumenGP)
```

### Load and Import Example Data

The package includes a small example dataset that can be used
to learn the workflow and test package functionality.

```{r}
files <- example_data()

raw_data <- read_ankom(
  files$ankom
)

metadata <- read_metadata(
  files$metadata
)
```

### Inspect ANKOM Data

The ANKOM RF file contains pressure measurements collected
during the incubation period.

```{r}
head(raw_data)
```

### Inspect Metadata

The metadata table contains bottle identifiers,
treatments, and replicate information.

```{r}
head(metadata)
```

### Validate metadata

```{r}
metadata <- validate_metadata(
  metadata
)
```

### Process ANKOM data

```{r}
gp <- process_ankom(
  raw_data,
  metadata,
  headspace_ml = 210,
  temperature_c = 39,
  zero_negative_pressure = TRUE
)
```

### Validate processed data

```{r}
gp <- validate_ankom(
  gp
)
```

---

## Example Workflow: Manual Gas Volume Data

```{r}
manual_volume <- data.frame(

  Bottle = c(
    1,1,1,
    2,2,2
  ),

  Treatment = c(
    "Control","Control","Control",
    "Corn","Corn","Corn"
  ),

  Time = c(
    0,4,8,
    0,4,8
  ),

  Gas = c(
    0,20,40,
    0,35,60
  )

)

gp_manual <- as_rumen_gp(
  data = manual_volume,
  head_col = "Bottle",
  treatment_col = "Treatment",
  time_col = "Time",
  gas_col = "Gas"
)

head(gp_manual)
```

---

## Example Workflow: Pressure Data

```{r}
manual_pressure <- data.frame(

  Bottle = rep(
    1,
    10
  ),

  Time = c(
    0,2,4,6,8,
    12,16,24,36,48
  ),

  PSI = c(
    0,
    0.2,
    0.5,
    0.8,
    1.2,
    1.8,
    2.5,
    3.2,
    4.0,
    4.5
  )

)

gp_pressure <- as_rumen_gp(
  data = manual_pressure,
  head_col = "Bottle",
  time_col = "Time",
  pressure_col = "PSI",
  pressure_unit = "psi",
  headspace_volume = 60
)

head(gp_pressure)
```

Pressure measurements are automatically converted
to gas volume using the same conversion framework
used by `process_ankom()`.

---

## Model Fitting

```{r}
groot_fit <- fit_groot(gp)

brody_fit <- fit_brody(gp)

gompertz_fit <- fit_gompertz(gp)
```

### Model summaries

```{r}
summary(groot_fit)
```

---

## Compare Models

```{r}
comparison <- compare_models(

  Groot = fit_groot(gp),

  Brody = fit_brody(gp),

  Gompertz = fit_gompertz(gp)

)

comparison
```

### Rank Models

```{r}
rank_models(
  comparison
)
```

---

## Treatment-Level Comparison

```{r}
treatment_comparison <-
  compare_models_by_treatment(

    Groot = fit_groot(gp),

    Brody = fit_brody(gp),

    Gompertz = fit_gompertz(gp)

  )

ranked_treatments <-
  rank_models_by_treatment(
    treatment_comparison
  )

best_model_by_treatment(
  ranked_treatments
)
```

---

## User-Defined Models

Researchers can fit their own kinetic equations
using `fit_custom()`.

```{r}
custom_fit <- fit_custom(

  data = gp,

  formula =
    Gas_mL ~
      A *
      (
        Time_h /
        (
          Time_h + K
        )
      ),

  start = list(
    A = 150,
    K = 10
  ),

  lower = c(
    A = 0,
    K = 0
  ),

  model_name = "Hyperbolic"

)

summary(custom_fit)
```

Compare custom and built-in models:

```{r}
compare_models(

  Groot = fit_groot(gp),

  Brody = fit_brody(gp),

  Hyperbolic = custom_fit

)
```

---

## Visualization

### Raw gas production

```{r, eval = FALSE}
plot_gp(
  gp,
  head = "1"
)
```

### Plot model fit

```{r, eval = FALSE}
plot_fit(
  groot_fit,
  head = "1"
)
```

### Plot residuals

```{r, eval = FALSE}
plot_residuals(
  groot_fit,
  head = "1"
)
```

### Treatment means

```{r, eval = FALSE}
plot_treatment_mean(
  Groot = fit_groot(gp),
  Brody = fit_brody(gp),
  treatment = "Control"
)
```

---

## Quality Control Workflow

```text
Import data
    ↓
Validate metadata
    ↓
Create rumen_gp object
    ↓
Fit models
    ↓
Flag problematic bottles
    ↓
Inspect residuals
    ↓
Document exclusions
    ↓
Refit models
    ↓
Compare models
```

Example:

```{r, eval = FALSE}
gp_clean <- exclude_heads(
  gp,
  heads = c("10"),
  reason = "Sensor malfunction"
)
```

---

## Implemented Models

### Built-in Models

- Brody
- Dual Logistic
- EXP0
- EXPL
- Gompertz
- Groot
- LE0
- LEL
- Logistic
- Mitscherlich
- Michaelis-Menten
- Orskov and McDonald

### Model Notes

The Groot and generalized Michaelis-Menten
models are mathematically equivalent.

Parameter correspondence:

- VF = A
- b = K
- k = c

---

## Major Features

### Data Import

- `process_ankom()`
- `as_rumen_gp()`

### Modeling

- 12 built-in models
- `fit_custom()`

### Diagnostics

- `summary()`
- `flag_model()`
- `exclude_heads()`

### Visualization

- `plot_gp()`
- `plot_fit()`
- `plot_residuals()`
- `plot_treatment_mean()`

### Model Comparison

- `compare_models()`
- `rank_models()`
- `compare_models_by_treatment()`
- `rank_models_by_treatment()`
- `best_model_by_treatment()`
- `model_win_frequency()`

---

### Completed

- ✅ ANKOM RF workflow
- ✅ Generic data import
- ✅ Pressure-based workflows
- ✅ Twelve kinetic models
- ✅ Custom model framework
- ✅ Model comparison framework
- ✅ Treatment-level model ranking
- ✅ Diagnostic workflows
- ✅ Visualization tools
- ✅ Five package vignettes
- ✅ pkgdown website
- ✅ Clean R CMD check

### Development Status

rumenGP is currently maintained and developed
as an open-source framework for rumen gas
production analysis. Future releases may include
additional modeling, reporting, and statistical
workflows.

---

## Citation

If you use **rumenGP** in research, please cite:

```text
Rodrigues, A. A. and Mantovani, H. C.

rumenGP:
An R package for rumen gas production
kinetic modeling, model comparison,
and visualization.
```

(Citation information will be updated as the package develops.)
